---
title: "A brief analysis of the Global Terrorism Database"
author: "Julian Hatwell"
date: "April 1st, 2016"
output: html_document
---

```{r prologue, results='hide', echo=FALSE}
knitr::opts_chunk$set(warning = FALSE
                      , message = FALSE
                      , echo = FALSE
                      , results = 'hide'
                      , fig.show='hide'
                      )
```

```{r setup}
require(readr)
require(MASS)
require(dplyr)
require(reshape2)
require(ggplot2)
require(vcd)
require(vcdExtra)
require(splines)
require(car)
require(countreg)

intextCols = c("pink3", "darkblue")
yrmnth.grid <- data.frame(nMonthsSinceJan1970 = 1:552)
series_base <- data.frame(n = rep(0,552), nMonthsSinceJan1970 = 1:552)
series_base$yrmnth <- as.character(1:552)
yr <- factor(floor((series_base$nMonthsSinceJan1970-1) / 12 + 1970))
pos.part <- function(x) {
  x <- ifelse(x < 0, 0, x)
  x
}
```

```{r getDataLocal, cache=TRUE}
gtd <- read_csv("gtd.csv")
```

```{r tidyData}
gtdwe <- gtd[gtd$region_txt == "Western Europe" &
               gtd$nkill > 0 & !(is.na(gtd$nkill))
             , c("eventid", "iyear", "imonth", "country_txt", "nkill", "gname")]

gtdwe <- within(gtdwe, {
  country_txt <- ifelse(country_txt %in% c("Italy", "France"
                                           , "Germany", "United Kingdom", "Spain")
                        , country_txt, "Other")
  imonth[eventid == 197400000001] <- 4 # Red Brigades Genoa 1974 Kidnapping
})
rm("gtd")
```

```{r basePlotData}
gtdfatals <- tapply(gtdwe$nkill, INDEX = list(gtdwe$iyear, gtdwe$country), FUN = sum, na.rm = TRUE)
gtdfatals <- data.frame(year = rownames(gtdfatals), gtdfatals)
gtdfatals <- melt(id.vars = "year"
                  , variable.name = "country"
                  , value.name = "fatalities"
                  , data = gtdfatals)
gtdfatals <- gtdfatals[!is.na(gtdfatals$fatalities),]
gtdfatals <- within(gtdfatals, {
  country <- factor(country, levels = c("Other", "United.Kingdom", "Italy", "Spain", "France", "Germany"))
  year <- factor(year, levels = 2015:1970)
  })
gtdfatals <- rbind(gtdfatals, list(2015, "France", 130))
```

```{r groupsData}
gtdgroups <- tapply(gtdwe$nkill
, INDEX = list(gtdwe$iyear, gtdwe$imonth, gtdwe$gname)
, FUN = sum, na.rm = TRUE)
gtdgroups <- data.frame(ftable(gtdgroups, row.vars = 1:3)) %>%
filter(!(is.na(Freq))) %>%
rename(fatalities = Freq
, tergroup = Var3
, month = Var2) %>%
mutate(year = factor(Var1, levels = 2015:1970))
gtdgroups <- within(gtdgroups, {
levels(tergroup) <- c(levels(tergroup), "Lockerbie"
, "Paris Attacks")
})
gtdgroups <- gtdgroups[, c("year", "month", "tergroup", "fatalities")]

gtdgroups[gtdgroups$year == 1988 &
gtdgroups$tergroup == "Unknown"
, "tergroup"] <- "Lockerbie"
gtdgroups <- rbind(gtdgroups, list(2015, 11, "Paris Attacks", 130))
gtdgroups$tergroup <- make.names(gtdgroups$tergroup)

#load("C:\\Dev\\Study\\R\\R_Basics\\tervars.RData")
load("D:\\github\\R_Basics\\tervars.RData")
tervars <- tervars %>% mutate(tergroup = make.names(tergroup)) 
gtdgroups <- left_join(gtdgroups, tervars) %>%
mutate(tergroup = factor(tergroup)
, International.Extremism = factor(intext, labels = c("No", "Yes")))
```

```{r numberOfIncidentsData}
gtdgroups <- within(gtdgroups, {
  yr <- as.numeric(as.character(year)) - 1970
  Year <- as.numeric(as.character(year))
  mnth <- as.numeric(as.vector(month))
  yrmnth <- yr * 12 + mnth
  ieCols <- ifelse(International.Extremism == "No"
                   , intextCols[1], intextCols[2])
})

nincm <- tapply(gtdgroups$International.Extremism, gtdgroups$yrmnth, function(x) 
{sum(1 * (x == "No")) })

nincm <- cbind(series_base, nincm = as.vector(nincm[series_base$yrmnth]))
nincm$nIncidents <- nincm$n + ifelse(!(is.na(nincm$nincm)), nincm$nincm, 0)
nincm$year <- yr

c(mean = mean(nincm$nIncidents), var = var(nincm$nIncidents),
                ratio = var(nincm$nIncidents) / mean(nincm$nIncidents))

yincm <- tapply(gtdgroups$International.Extremism, gtdgroups$yrmnth, function(x) 
{sum(1 * (x == "Yes")) })

yincm <- cbind(series_base, yincm = as.vector(yincm[series_base$yrmnth]))
yincm$nIncidents <- yincm$n + ifelse(!(is.na(yincm$yincm)), yincm$yincm, 0)
yincm$year <- yr

c(mean = mean(yincm$nIncidents), var = var(yincm$nIncidents),
                ratio = var(yincm$nIncidents) / mean(yincm$nIncidents))
```

```{r fatalitiesData}
fatm <- t(tapply(gtdgroups$fatalities, INDEX = list(gtdgroups$International.Extremism, gtdgroups$yrmnth), FUN = sum))

nfatm <- fatm[, "No"]
nfatm <- cbind(series_base, nfatm = as.vector(nfatm[series_base$yrmnth]))
nfatm$fatalities <- nfatm$n + ifelse(!(is.na(nfatm$nfatm)), nfatm$nfatm, 0)
nfatm$year <- yr

yfatm <- fatm[, "Yes"]
yfatm <- cbind(series_base, yfatm = as.vector(yfatm[series_base$yrmnth]))
yfatm$fatalities <- yfatm$n + ifelse(!(is.na(yfatm$yfatm)), yfatm$yfatm, 0)
yfatm$year <- yr
```

```{r nfatPoisFitBad}
nfatc.pois <- goodfit(table(nfatm$fatalities), type = "poisson")
sum.nfatc.pois <- summary(nfatc.pois)
unlist(nfatc.pois$par)
plot(nfatc.pois, shade = TRUE
     , main = "poisson model"
     , xlab = "Number of incidents\nin a given month")
distplot(table(log(nfatm$fatalities+1)), lambda = unlist(nfatc.pois$par)[1]
        , xlab = "Number of incidents\nin a given month")
```

```{r nfatPoisFit}
nfatc.pois <- goodfit(table(log(nfatm$fatalities+1)), type = "poisson")
sum.nfatc.pois <- summary(nfatc.pois)
unlist(nfatc.pois$par)
plot(nfatc.pois, shade = TRUE
     , main = "poisson model"
     , xlab = "log(fatalities) in a given month")
distplot(table(log(nfatm$fatalities+1)), lambda = unlist(nfatc.pois$par)[1]
        , xlab = "log(fatalities) in a given month")
```

```{r nfatNbinFit}
nfatc.nbin <- goodfit(table(log(nfatm$fatalities+1)), type = "nbinom")
sum.nfatc.nbin <- summary(nfatc.nbin)
unlist(nfatc.nbin$par)
plot(nfatc.nbin, shade = TRUE
     , main = "negative binomial model"
     , xlab = "log(fatalities) in a given month")
distplot(table(log(nfatm$fatalities+1)), type = "nbinomial", size = unlist(nfatc.nbin$par)[1]
        , xlab = "log(fatalities) in a given month")
```

```{r yfatPoisFit}
yfatc.pois <- goodfit(table(log(yfatm$fatalities+1)), type = "poisson")
sum.yfatc.pois <- summary(yfatc.pois)
unlist(yfatc.pois$par)
plot(yfatc.pois, shade = TRUE
     , main = "poisson model"
     , xlab = "log(fatalities) in a given month")
distplot(table(log(yfatm$fatalities+1)), lambda = unlist(yfatc.pois$par)[1]
        , xlab = "log(fatalities) in a given month")
```

```{r yfatNbinFit}
yfatc.nbin <- goodfit(table(log(yfatm$fatalities+1)), type = "nbinom")
sum.yfatc.nbin <- summary(yfatc.nbin)
unlist(yfatc.nbin$par)
plot(yfatc.nbin, shade = TRUE
     , main = "negative binomial model"
     , xlab = "log(fatalities) in a given month")
distplot(table(log(yfatm$fatalities+1)), type = "nbinomial", size = unlist(yfatc.nbin$par)[1]
        , xlab = "log(fatalities) in a given month")
```

```{r nfatNbinomGLM}
gtf.glm <- glm(fatalities~nMonthsSinceJan1970, data = nfatm
                , family = negative.binomial(theta = unlist(nfatc.nbin$par)[1], link=log))

gtf.nlm <- glm(fatalities~bs(nMonthsSinceJan1970
                             , knots = c(60, 150, 240, 420))
               , data = nfatm
               , family = negative.binomial(theta = unlist(nfatc.nbin$par)[1], link=log))

gtf.hur <- hurdle(fatalities~nMonthsSinceJan1970
               , data = nfatm, dist = "negbin", zero.dist = "negbin")
```

```{r yfatNbinomGLM}
gtf.glm2 <- glm(fatalities~nMonthsSinceJan1970, data = yfatm
                , family = negative.binomial(theta = unlist(yfatc.nbin$par)[1], link=log))

gtf.nlm2 <- glm(fatalities~ns(nMonthsSinceJan1970, knots = c(150, 300))
                , data = yfatm
                , family = negative.binomial(theta = unlist(yfatc.nbin$par)[1], link=log))

gtf.hur2 <- hurdle(fatalities~nMonthsSinceJan1970
               , data = yfatm, dist = "negbin", zero.dist = "negbin"
               , link = "log")
```

## Introduction

This post is inspired by a recent [article in the Huffington Post](http://www.huffingtonpost.co.uk/2015/11/28/islamic-state-terrorism-threat_n_8670458.html) which contained a barchart (see appendices) of fatalities caused by terrorism in Western Europe over the years since 1970.

The original source of the data is the [Global Terrorism Database](https://www.start.umd.edu/gtd/), which is freely available to download. It's fairly trivial to reconstruct the graph from this base data, as I show here. The original is in the appendices:

```{r basePlot, fig.show='show'}
gg <- ggplot(data = gtdfatals
             , aes(x = year, y = fatalities, fill = country))
gg + geom_bar(stat = "identity", colour = "lightgrey") + 
  scale_fill_manual(values=c("darkgrey", "steelblue", "darkgreen", "gold", "red", "lightblue")) +
  coord_flip() + theme_bw() + scale_x_discrete(drop = FALSE) +
  theme(legend.position="top", axis.text.y = element_text(size=6)) +
  labs(title = "Victims of Terrorist Attacks in Western Europe", y = "Number of fatalities", x = "Year")
```

On initial visual assessment, the eye is drawn to the mountain of bars representing the 70's and 80's and the relative sparsity from the mid-90's onwards. After 2001 there are only the 4 notable peaks in 2004 (Madrid), 2005 (London), 2011 (Norway) and 2015 (Paris). One can immediately gain a sense that the danger from terrorism from 1970 to the end of the 1990's was much greater than it is now.

In my opinion, the graph is a fairly successful attempt to demonstrate that our current perception and fear of global terrorism is caused by something other than factual measures such as the number of deaths (or number of incidents). Our collective perception may be more a symptom of the tactics of terror or simply the dominant socio-political narrative. This unease we feel causes us to find rationales for ignoring the facts of our recent history and to allow ourselves to believe that no time is as dangerous as now.

My reposting of the original graphic on Facebook drew the attention of one particular friend who asserted that the graph wasn't making a valid comparison to current Islamist terrorism because it "includes Northern Ireland in the 1970s and ETA in the 1980s which is not quite the same."

He went on to say "ETA and IRA conflicts were localised not global, secondly, it could be argued that they were liberation movements in search of a political solution and as a result they had the sympathy of parties on the Left. Corbyn and Livingstone supported Sinn Fein but not Al Qaeda or ISIS."

Essentially these comments serve well to emphasise my point. Can we say that people killed in Ireland by the IRA are less dead than people killed in Paris by Islamic extremists? Or were bombs planted by the IRA less dangerous because Jeremy Corbyn liaised with Gerry Adams when he was otherwise shut out of the political discourse? I think not and neither does the [Global Terrorism Database](https://www.start.umd.edu/gtd/) make any such distinction. 

This inspired me to do my own analysis of the [Global Terrorism Database](https://www.start.umd.edu/gtd/) to see what (objective) stories the data has to tell and help to cut through some of the political noise and cognitive bias that surrounds such an emotive topic.

## Exploring the data

To examine more closely the different trends in localised terrorism compared to international extremism  it's necessary to determine which are the local and international terror groups in the subset of the data pertaining to Western Europe and incidents with 1 or more fatality. 

The [Global Terrorism Database](https://www.start.umd.edu/gtd/) does not make this distinction explicit. I had to go through the list myself and label all the groups that I could be sure about (e.g. Basque separatists and Irish Republicans, compared to Secret Organization of al Qa'ida in Europe). I share my categories in the appendices.

My first question, of all the deaths from terrorism recorded in the database since 1970 (plus 130 from Paris 2015), how many are from International terrorism?

```{r AllDeaths, fig.show='show', fig.height=3}
DeathsAll <- with(gtdgroups, tapply(fatalities, International.Extremism, sum))
barplot(DeathsAll, col = intextCols
        , main = "Fatalities from Terror Attacks in Western Europe\n1970-2015"
        , xlab = "Action by International Extremists?")
text(c(0.7, 1.9), 5600, DeathsAll, xpd = TRUE)
```

Judging by the numbers here, `r round(DeathsAll[1]/DeathsAll[2],1)` times as many deaths were caused by localised terrorism between 1970 and 2015. 

Given many of these localised groups have been much quieter since September 11th bombings of 2001 and we've had the Good Friday Agreement (GFA) working well since December 1999, it might help to look at the balance since the GFA:
  
```{r PostGFADeats, fig.show='show', fig.height=3}
DeathsGFA <- with(gtdgroups[as.numeric(as.character(gtdgroups$year)) >= 2000,], tapply(fatalities, International.Extremism, sum))
barplot(DeathsGFA, col = intextCols
        , main = "Fatalities from Terror Attacks in Western Europe\n2001-2015"
        , xlab = "Action by International Extremists?"
        , ylim = c(0, max(DeathsAll)))
text(c(0.7, 1.9), 1000, DeathsGFA, xpd = TRUE)
```

The balance has shifted to `r round(DeathsGFA[2]/DeathsGFA[1],1)` times as many deaths from international terrorists as localised terrorism. The international threat must feel more current, but is still altogether out of balance given the totals. Take a look at the proportions:

```{r, mosaicPrePostGFA, fig.show='show', fig.height=4, fig.width=4.5}
gtdgroups <- within(gtdgroups, {
  GFA <- factor(ifelse(as.numeric(as.character(year)) >= 2000, "Post", "Pre"), levels = c("Pre", "Post"))
})

DeathsMosaic <- with(gtdgroups, tapply(fatalities, INDEX = list(GFA, International.Extremism), sum))
largs <- list(set_varnames = c(B = "Action by International Extremists",
                               A = "Relative to GFA")
              , cex = 0.9)
mosaic(DeathsMosaic
      , gp = gpar(col = intextCols, lwd = c(5,2,2,5), lty = c(1,2,2,1))
      , labeling_args = largs
      , labeling = labeling_values
      , legend = FALSE
      , keep_aspect_ratio = FALSE
      , main = "Fatalities from terrorist attacks 1970-2015"
      , main_gp = gpar(fontsize = 12))

mosaic(round(prop.table(DeathsMosaic),2) * 100
      , gp = gpar(col = intextCols, lwd = c(5,2,2,5), lty = c(1,2,2,1))
      , labeling_args = largs
      , labeling = labeling_values
      , legend = FALSE
      , keep_aspect_ratio = FALSE
      , main = "Fatalities (%) from terrorist attacks 1970-2015"
      , main_gp = gpar(fontsize = 12))
```

The dashed outlines show where numbers are lower than expected given the overall proportions. In short, the weight of fatalities from international terrorist attacks falls disproportionately after the GFA. This is no surprise. 

What's important is this; Based on an assumption that things are more dangerous now, in the 15 years after the GFA you'd expect to see at least half as many fatalities as in the 30 previous years. The split is 90% - 10%

Here's the original graph with the local vs international split highlighted:

```{r intextPlot, fig.show='show'}
gg <- ggplot(data = gtdgroups
, aes(x = year, y = fatalities)) +
scale_fill_manual(values=intextCols)
gg + geom_bar(stat = "identity", aes(fill = International.Extremism)) +
coord_flip() + theme_bw() + scale_x_discrete(drop = FALSE) +
  theme(legend.position="top", axis.text.y  = element_text(size=6)) +
labs(title = "Victims of Terrorist Attacks in Western Europe", y = "Number of fatalities", x = "Year")
```

One should conclude that those `r DeathsGFA[2]` deaths caused by global terrorists in the years since the GFA are almost entirely accounted for in Madrid and London bombings and the Paris attacks. Clearly these are very deadly incidents. However, it is now easy to pick out some equally large events in the data going back to the 1970's and 1980's. It would be useful see if the lethality of the attacks has changed over time.

```{r AvgDeaths, fig.show='show', fig.height=3}
DeathsAllAve <- with(gtdgroups, tapply(fatalities, International.Extremism, mean))
DeathsGFAAve <- with(gtdgroups[as.numeric(as.character(gtdgroups$year)) >= 2000,], tapply(fatalities, International.Extremism, mean))

barplot(c(DeathsAllAve,DeathsGFAAve), col = intextCols
        , main = "Average Fatalities per Terrorist Incident in Western Europe"
        , xlab = "Action by International Extremists?")
abline(v = 2.5, lwd = 3)
text(c(0.6, 1.85, 3.1, 4.3)
     , 102
     , round(c(DeathsAllAve,DeathsGFAAve),0)
     , xpd = TRUE)
text(2.4, 40, "All Count\n1970-2015", pos = 2)
text(2.6, 40, "Since GFA\n2001-2015", pos = 4)
```

Looking at average death toll alone seems to indicate that when they strike, the international terrorists are deadlier. They appear to have become vastly more deadly per incident in the last 15 years. However, this simplistic explanation does not tell the whole story, as we shall see.

The following two graphs restate the same data as counts of numbers of incidents that result in fatalities:

```{r intextCountPlot, fig.show='show'}
gg2 <- ggplot(data = gtdgroups
, aes(x = year)) +
scale_fill_manual(values=intextCols)
gg2 + geom_bar(stat = "count", aes(fill = International.Extremism)) +
scale_x_discrete(drop = FALSE) +
coord_flip() + theme_bw() + theme(legend.position="top"
, axis.text.y = element_text(size=6)) +
labs(title = "Number of terrorist incidents
resulting in fatality in Western Europe"
, y = "Number of incidents", x = "Year")
```

Here, the localised terror groups absolutely dominate the graph. It's also obvious that there was a far greater, more constant and continuous threat in the past. All incidents are rapidly diminishing in number to the present day.

Looking at the international incidents alone, we can see that they follow the same trend, decreasing in number since the 1970's-80's:

```{r extCountPlot, fig.show='show'}
gg3 <- ggplot(data = gtdgroups[gtdgroups$International.Extremism == "Yes",]
, aes(x = year))
gg3 + geom_bar(stat = "count", fill = "darkblue") +
scale_x_discrete(drop = FALSE) +
coord_flip() + theme_bw() + theme(legend.position="top"
, axis.text.y = element_text(size=6)) +
labs(title = "Number of terrorist incidents
committed by international extremists
resulting in fatality in Western Europe"
, y = "Number of incidents", x = "Year")
```

This is the clearest evidence that not only are global terror attacks are much rarer than localised groups over period 1970-2015, they have also become far more rare in recent years since the 1990's.

So back to the question of whether the attacks have become deadlier in recent years. If that is the case then surely we do have as much to fear from larger scale attacks, even if they are less frequent.

Firstly, a look at the localised groups:

```{r nfatNbinomGLMplot, fig.show='show', fig.height=4, fig.width=4}
gtf.glm.preds <- predict(gtf.glm, newdata = yrmnth.grid, se = TRUE)
gtf.hur.preds <- data.frame(fit = predict(gtf.hur, newdata = yrmnth.grid, type = "response"))

gtf.hur.preds <- within(gtf.hur.preds, {
  ci.upr <- fit + 1.96 * gtf.glm.preds$se.fit
  ci.lwr <- fit - 1.96 * gtf.glm.preds$se.fit
})
gtf.hur.preds <- lapply(gtf.hur.preds, pos.part)

gtf.nlm.preds <- predict(gtf.nlm, newdata = yrmnth.grid, type = "response", se = TRUE)

gtf.nlm.preds <- within(gtf.nlm.preds, {
  ci.upr <- fit + 1.96 * se.fit
  ci.lwr <- fit - 1.96 * se.fit
})
gtf.nlm.preds <- lapply(gtf.nlm.preds, pos.part)

plot(log(fatalities+1)~nMonthsSinceJan1970
     , data = nfatm#, xlim = c(0,600)
     , xaxt = "n"
     , col = intextCols[1]
     , pch = 20, cex =0.5
     , xlab = "Year"
     , ylab = "log(Fatalities) in a given month"
     , main = "Fatalities from terrorist actions\n (local, non-international)")

with(gtf.nlm.preds
     , polygon(c(1:552,552:1)
               , log(c(ci.upr, rev(ci.lwr))+1)
               , col=rgb(0,0.8,0.5,0.1), border=NA))
lines(1:552, log(gtf.nlm.preds$fit+1), col = "aquamarine1", lwd = 3)

with(gtf.hur.preds
     , polygon(c(1:552,552:1)
               , log(c(ci.upr, rev(ci.lwr))+1)
               , col=rgb(1,0.3,0.4,0.15), border=NA))
lines(1:552, log(gtf.hur.preds$fit+1), col = intextCols[1], lwd = 3)

axis(side = 1, at = seq(0, 600, 60), labels = seq(0, 600, 60)/12 +1970)
points(c(128,499), log(c(85,77)), pch = c(4,1), col = "red", cex = 1.5)
text(c(128,499), log(c(85,77)), pos = c(4,1), c("Bologna, 1980", "Norway, 2011"), cex = 0.6)

legend("bottomleft"
       , inset = c(0.15, 0)
       , legend = c("log(fatalities)"
                   , "hurdle fit"
                   , "basis function")
       , cex = 0.6
       , col = c(intextCols[1], intextCols[1], "aquamarine2")
       , pch = c(19, NA, NA)
       , lty = c(NA, 1, 1)
       , lwd = c(NA, 2, 2))

plot(log(fatalities)~year, nfatm, col = intextCols[1]
     , main = "Distribution of fatalities\n(local, non-international)")
points(42, log(77), col = "red", cex = 1.5)
text(42, log(77), pos = 2, "Norway, 2011", cex = 0.6)
```

The points are aggregated such that August 1980 shows up as 137 fatalities. In fact the most lethal non-International terrorist attack in Western Europe was the ["Strage di Bologna"](https://en.wikipedia.org/wiki/Bologna_massacre)  which killed 85, allegedly carried out by the neo-fascist Nuclei Armati Rivoluzionari. A mark has been added to indicate where this would lie on the plot as a standalone point. Note how this compares to the second single most lethal attack, carried out by Anders Breivik in Norway (also indicated). The official explanation is that he was an individual unlikely to have been linked to any group.

For the localised terrorist attacks, the steady downward trend obvious and this echoes what we know to be true from history; In Western Europe, terrorism from local sources (Nationalism, Republicanism, Revolutionary Communism, Fascism) was very prevalent in the 1970's and 1980's. From this peak it has steadily declined and all but disappeared today.

I continue along the same lines for the international extremist terror threat:

```{r yfatNbinomGLMplot, fig.show='show', fig.height=4, fig.width=4}
gtf.glm.preds <- predict(gtf.glm2, newdata = yrmnth.grid, se = TRUE)
gtf.hur.preds <- data.frame(fit = predict(gtf.hur2, newdata = yrmnth.grid, type = "response"))

gtf.hur.preds <- within(gtf.hur.preds, {
  ci.upr <- fit + 1.96 * gtf.glm.preds$se.fit
  ci.lwr <- fit - 1.96 * gtf.glm.preds$se.fit
})
gtf.hur.preds <- lapply(gtf.hur.preds, pos.part)

gtf.nlm.preds <- predict(gtf.nlm2, newdata = yrmnth.grid
                         , type = "response"
                         , se = TRUE)

gtf.nlm.preds <- within(gtf.nlm.preds, {
  ci.upr <- fit + 1.96 * se.fit
  ci.lwr <- fit - 1.96 * se.fit
})
gtf.nlm.preds <- lapply(gtf.nlm.preds, pos.part)

plot(log(fatalities)~nMonthsSinceJan1970
     , data = yfatm
     , xaxt = "n"
     , col = intextCols[2]
     , pch = 20, cex =0.5
     , xlab = "Year"
     , ylab = "log(Fatalities) in a given month"
     , main = "Fatalities from terrorist actions\n (international in Western Europe)")

axis(side = 1, at = seq(0, 600, 60), labels = seq(0, 600, 60)/12 +1970)

with(gtf.nlm.preds
     , polygon(c(1:552,552:1)
              , log(c(ci.upr, rev(ci.lwr))+1)
              , col=rgb(0,0.8,0.5,0.1), border=NA))               
lines(1:552, log(gtf.nlm.preds$fit+1), col = "aquamarine2", lwd = 3)

with(gtf.hur.preds
     , polygon(c(1:552,552:1)
              , log(c(ci.upr, rev(ci.lwr))+1)
              , col=rgb(0.2,0.3,1,0.15), border=NA))
lines(1:552, log(gtf.hur.preds$fit+1), col = intextCols[2], lwd = 3)

legend("bottomright", legend = c("log(fatalities)"
                              , "hurdle fit"
                              , "nspline")
       , inset = c(0.1, 0.4)
       , cex = 0.6
       , col = c(intextCols[2], intextCols[2], "aquamarine3")
       , pch = c(19, NA, NA)
       , lty = c(NA, 1, 1)
       , lwd = c(NA, 2, 2))

points(c(411,427,551)
       , log(c(191, 56, 130))
       , col = "red", cex = 1.5)
text(c(411,427,551)
     , log(c(191, 56, 130))
     , pos = c(2,4,2)
     , cex = 0.6
     , c("Madrid, 2004"
         , "London 2005"
         , "Paris 2015"))

points(c(2, 191,57,228)
       , log(c(55, 60, 88, 270))
       , col = "red", cex = 1.5)
text(c(2, 191,57,228), log(c(55, 60, 88, 270))
     , pos = c(4, 4,4,2)
     , cex = 0.6
     , c("\nSwissair 330"
         , "EgyptAir 648"
         , "TWA 841"
         , "Pan Am 103"))

plot(log(fatalities)~year, yfatm, col = intextCols[2]
     , main = "Distribution of fatalities\n(international in Western Europe)")
points(c(35,36,46)
       , log(c(191, 56, 130))
       , col = "red", cex = 1.5)
text(c(35,36,46)
     , log(c(191, 56, 130))
     , pos = c(2,4,2)
     , cex = 0.6
     , c("Madrid, 2004"
         , "London 2005"
         , "Paris 2015"))
points(c(1,16,5,19)
       , log(c(55, 60, 88, 270))
       , col = "red", cex = 1.5)
text(c(1,16,5,19), log(c(55, 60, 88, 270))
     , pos = c(4, 4,4,2)
     , cex = 0.6
     , c("\nSwissair 330"
         , "EgyptAir 648"
         , "TWA 841"
         , "Pan Am 103"))
```

Here we see a mostly flat trend in the in the 1970's to 90's followed by a steady decline. It's the same trend as the non-International group except it starts from a lower base. 

There is a majority of low level attacks with a small number of very serious incidents involving aeroplane bombings in the 70's and 80's, the public transport suicide bombings in Madrid and London ('04 and '05) and the death cell that attacked Paris last year.

The four catastrophic flight bombings easily rival the death tolls of more recent attacks. If the fatalities are trending downwards for both groups, why was there an apparent 10-fold increase in the average number of fatalities per incident? 

The problem is that it's a somewhat crude measure. It takes no account of all the months where there were no fatal attacks. Consider instead the fatalities per month over the whole period. This acts as a weighting against very rare events:

```{r AvgDeathsMonth, fig.show='show', fig.height=4}
DeathsAveMnthAll <- c("No" = round(mean(nfatm$fatalities,0))
                       , "Yes" = round(mean(yfatm$fatalities),0))
DeathsAveMnthGFA <- c("No" = round(mean(nfatm$fatalities[nfatm$nMonthsSinceJan1970 > 372],0))
                       , "Yes" = round(mean(yfatm$fatalities[yfatm$nMonthsSinceJan1970 > 372]),0))

barplot(c(DeathsAveMnthAll,DeathsAveMnthGFA), col = intextCols
        , main = "Average Terrorist Fatalities per Month in Western Europe"
        , xlab = "Action by International Extremists?")
abline(v = 2.5, lwd = 3)
text(c(0.7, 1.9, 3.1, 4.3)
     , 9.5
     , round(c(DeathsAveMnthAll,DeathsAveMnthGFA),0)
     , xpd = TRUE)
text(2.4, 4, "All Count\n1970-2015", pos = 2)
text(2.6, 4, "Since GFA\n2001-2015", pos = 4)
```

Accounting for all the months when there were no incidents, it's clear that the localised terrorists were the more deadly group in the period 1970-2000, by the sheer number of incidents even though international terrorists carried out a number of airplane bombing atrocities, these were no more than a handful in two decades.

The threat from international terrorists, if measured in a rate of fatalities over a given time period, has remained constant and low and has never come close to the danger we faced in the period 1970 - 2000 from localised terrorism representing a mix of sectarian, separatist, revoltionary communist and fascist ideologies in different countries at the time.

This analysis deals with the situation in Western Europe only and the conclusion one has to draw is that we're much safer now that we were in the period 1970-2000. To say that those atrocities were somehow different because they were restricted to places like Northern Ireland and Spain is to deny the horror and tragedy felt by the people who lived in those places at the time. One could argue that today's threat is different because the a huge majority of the deaths that result from recent atrocities is restricted to the Countries such as Iraq and Syria.

## Appendices

This was purely a theoretical exercise not intended to be taken as a prediction or forecast. There are many volatile factors that this picture could change dramatically and upredictably at any time. I'm looking at trends over time and not taking any influential factors in to consideration.

The analysis is done. If anyone is interested in the workings, read on.

### Original Chart

The original chart from the [article in the Huffington Post](http://www.huffingtonpost.co.uk/2015/11/28/islamic-state-terrorism-threat_n_8670458.html)

![barchart of fatalities][1]

[1]: GTD_files\\figure-html\\o-TERRORISM-570.jpg "Victims of Terrorist Attacks in Western Europe"

### Data Corrections

After downloading the freely available [Global Terrorism Database](https://www.start.umd.edu/gtd/) I found that it contains data only up to 2014. Huffington Post's statistical consultant probably added 2015-16 from other sources, although they don't tell us that. I've added a value of [130](https://en.wikipedia.org/wiki/November_2015_Paris_attacks#Casualties) fatalities in France for 2015. 

There is an entry for the Red Brigades 1974 hostage taking in Genoa with 1 killed which has month set as zero. [I have updated this to April](https://web.stanford.edu/group/mappingmilitants/cgi-bin/groups/view/77)

### Categorisation of the terrorist groups

The catogories applied to the various terror groups. Lockerbie was listed under "Unknown" but was easy to single out and label. Paris Attacks are also individually marked as this point has been added in. Any advice on making it more accurate is most welcome:

```{r cats, results='show'}
cats <- unique(gtdgroups[, c("tergroup", "International.Extremism")])
cats[,1] <- abbreviate(as.character(cats[,1]), minlength = 40)
cats
```

### The plot trend lines

The main trend lines were added using a linear hurdle model. Traditional linear models would be dragged downwards on the Y axis by the prevalence of zeros in the data, giving too low an estimate. The hurdle is robust to this type of data.

In the non-international plot, a basis function is used with elbows (knots) that follow the non-linear ups and downs in historical trends.

To make the fairest possible comparison in the international groups plot, the same basis function above was considered. However, it was found to be far too sensitive to outliers. The model used here is a natural spline which is a lot smoother. Still, it is less stable than the hurdle model because of the spread
between all the months with zero fatalities recorded and the number of points with very high recorded fatalities. This is reflected in the very large confidence bands and the uplift at the right extremity where it tries to head towards the outliers.

### Finding the best fit

For the scatter plots, finding the right models to add trend lines was not a traditional linear regression task as the data are not iid. 

I'm really looking at a time series where it is highly likely (as certain as can be) that near neighbours in the data will be correlated. Whatever are the factors that lead to terrorism one month are probably just as true in the previous and next months, but may very more from year to year, and most as the decades go by.

There are also many months where the number of incidents, and consequently fatalities, is zero. It's not a smooth distribution.

The challenge was to find a robust smoothing line that reveals trends in the count data where there is serious non-linearity and dramatic zero inflation.

I started with the assumption that number of fatalities in a given month follows standard models for count data, but initially struggled with it as can be seen from the first attempt to perform a poisson goodness of fit:

```{r nfatPoisFitBad, fig.show='show', fig.height=4, fig.width=4}
```

What's happening here? The data are very skewed and my obvious next step is to operate on the log transformed data (+1 for the zeros). When thinking about why that is, it makes more sense to think of the incidents themselves are the count data, while the number of deaths for each incident is really some level of intensity. This might even be considered to be quantitative, and certainly leaves me thinking that I'm actually trying to model a multivariate response of frequency and intensity. That's going to take quite a bit of brain power and will require a lot more explanatory variables to solve it. I put this problem aside for another day. 

Obviously I can start with the log transform to tame the high skew and dispersion.

```{r nfatPoisFit, fig.show='show', fig.height=4, fig.width=4}
```

Here in the poisson model I can see evidence of zero inflation. This means there are many more times when zero deaths are counted in a given month than the standard poisson model allows for.

```{r nfatNbinFit, fig.show='show', fig.height=4, fig.width=4}
```

These are a lot better with much less evidence of zero inflation. However, there's still a real lack of fit at $e^2$ and above. I want to find a model suitable for this situation.

I go through the same process with the fatalities data from international terror incidents.

```{r yfatPoisFit, fig.show='show', fig.height=4, fig.width=4}
```

Here there is huge zero inflation compared to what the model expects. Of course this stands to reason as there are many months out of the time period when there are no fatal incidents.

```{r yfatNbinFit, fig.show='show', fig.height=4, fig.width=4}
```

Again, the negative binomial is a far better, albeit rather imperfect fit. This can be corrected by means of a zero inflated or hurdle model:

```{r nfatRootFits, fig.show='show', fig.height=4, fig.width=3}
gtn_hnb <- hurdle(fatalities~nMonthsSinceJan1970
                  , data = nfatm, dist = "negbin")
gtn_znb <- zeroinfl(fatalities~nMonthsSinceJan1970
                  , data = nfatm, dist = "negbin")

countreg::rootogram(gtf.glm, max = 50, main = "Negative Binomial"
                    , xlab = "fatalities (non-international)")
countreg::rootogram(gtn_hnb, max = 50, main = "Hurdle Negative Binomial"
                    , xlab = "fatalities (non-international)")
countreg::rootogram(gtn_znb, max = 50
                    , main = "Zero-inflated Negative Binomial"
                    , xlab = "fatalities (non-international)")
```

```{r yfatRootFits, fig.show='show', fig.height=4, fig.width=3}
gty_hnb <- hurdle(fatalities~nMonthsSinceJan1970
                  , data = yfatm, dist = "negbin")
gty_znb <- zeroinfl(fatalities~nMonthsSinceJan1970
                  , data = yfatm, dist = "negbin")

countreg::rootogram(gtf.glm2, max = 50, main = "Negative Binomial"
                    , xlab = "fatalities (international)")
countreg::rootogram(gty_hnb, max = 50, main = "Hurdle Negative Binomial"
                    , xlab = "fatalities (international)")
countreg::rootogram(gty_znb, max = 50
                    , main = "Zero-inflated Negative Binomial"
                    , xlab = "fatalities (international)")
```

There's not much to choose between the latter two in each case. Hurdle seems more intuitive because all the zeros are structural. The data has already been filtered to include only incidents where there was 1 or more fatality, so all the zeros for fatalities come from the months without incident.